Expression QTL hotspot dissection with mediation analysis and pleiotropy testing

### 2020-03-25 analysis/paper/Rmd/hotspot-one-chromosome-TIMBR-Chr2.Rmd
set up for using child processes with TIMBR analyses

```{r, define_local_annot}
local_annot <- lod_peaks %>%
  filter(chrom == hot_chr, pos <= hot_upper, pos >= hot_lower) %>%
  left_join(y = annots, by = c("annot.id" = "gene_id")) %>%
  filter(!duplicated(annot.id)) %>%
  filter(lod >= ult) %>%
  mutate(local_tx = chr == hot_chr) %>%
  filter(local_tx)
```

```{r, define_nonlocal_annot}
nonlocal_annot <- lod_peaks %>%
  filter(chrom == hot_chr, pos <= hot_upper, pos >= hot_lower) %>%
  inner_join(y = annots, by = c("annot.id" = "gene_id")) %>%
  filter(!duplicated(annot.id)) %>%
  filter(lod >= 7.18) %>%
  mutate(local_tx = chr == hot_chr) %>%
  filter(!local_tx) %>%
  filter(!is.na(hotspot))
```

```{r, local_annot2}
local_annot2 <- local_annot %>%
  filter(middle <= hot_upper, middle >= hot_lower) %>% 
  arrange(desc(lod))
```

@@

```{r, hot_annot-hot_chr}
hot_annot <- local_annot2 %>%
  bind_rows(nonlocal_annot) %>%
  dplyr::select(annot.id, symbol)
```

@@

>>>>

```{r, define_local_annot}
local_annot <- lod_peaks %>%
  dplyr::filter(chrom == hot_chr, pos <= hot_upper, pos >= hot_lower) %>%
  dplyr::left_join(y = annots, by = c("annot.id" = "gene_id")) %>%
  dplyr::filter(!duplicated(annot.id)) %>%
  dplyr::filter(lod >= ult) %>%
  dplyr::mutate(local_tx = chr == hot_chr) %>%
  dplyr::filter(local_tx)
```

```{r, define_nonlocal_annot}
nonlocal_annot <- lod_peaks %>%
  dplyr::filter(chrom == hot_chr, pos <= hot_upper, pos >= hot_lower) %>%
  dplyr::inner_join(y = annots, by = c("annot.id" = "gene_id")) %>%
  dplyr::filter(!duplicated(annot.id)) %>%
  dplyr::filter(lod >= 7.18) %>%
  dplyr::mutate(local_tx = chr == hot_chr) %>%
  dplyr::filter(!local_tx) %>%
  dplyr::filter(!is.na(hotspot))
```

```{r, local_annot2}
local_annot2 <- local_annot %>%
  dplyr::filter(middle <= hot_upper, middle >= hot_lower) %>% 
  dplyr::arrange(desc(lod))
```

@@

```{r, hot_annot-hot_chr}
hot_annot <- local_annot2 %>%
  dplyr::bind_rows(nonlocal_annot) %>%
  dplyr::select(annot.id, symbol)
```

@@

```{r}
t_out <- readRDS(here::here("analysis", "data", "derived_data", "timbr-chr2-10ksamples.rds")) %>%
  dplyr::rename(timbr = t_out)
```

### 2020-03-25 analysis/paper/Rmd/hotspot-one-chromosome-timbr.Rmd
set up for using child processes with TIMBR analyses

```{r, readRDS-{{hot_chr}}}
pleio_mixed_list <- readRDS(here::here("analysis", "data", "derived_data", paste0("Chr", {{hot_chr}}, "_pleio.rds")))
pleio_local_list <- readRDS(here::here("analysis", "data", "derived_data", paste0("Chr", {{hot_chr}}, "_pleio_local.rds")))
pleio_nonlocal_list <- readRDS(here::here("analysis", "data", "derived_data", paste0("Chr", {{hot_chr}}, "_pleio_nonlocal.rds")))
med_list <- readRDS(here::here("analysis", "data", "derived_data", paste0("Chr", {{hot_chr}}, "_mediations.rds")))
```

@@

```{r, med_tib-{{hot_chr}}, message=FALSE, warning = FALSE}
med_tib <- med_list %>%
  purrr::map(bind_rows) %>%
  bind_rows() %>%
  as_tibble() %>%
  rename(lod_without_med = lod_no_med, lod_with_med = lod_med) %>%
  mutate(lod_diff = lod_without_med - lod_with_med, lod_diff_proportion = lod_diff / lod_without_med) %>%
  left_join(local_annot2, by = c("local_gene_id" = "annot.id")) %>%
  left_join(nonlocal_annot, by = c("nonlocal_gene_id" = "annot.id")) %>%
  dplyr::select(symbol.x, symbol.y, lod_without_med, lod_with_med, lod_diff, lod_diff_proportion, local_gene_id, nonlocal_gene_id) %>%
  rename(local_symbol = symbol.x, nonlocal_symbol = symbol.y) %>%
  filter(!is.na(local_symbol))
```

```{r, hot_annot-{{hot_chr}}}
hot_annot <- local_annot2 %>%
  bind_rows(nonlocal_annot) %>%
  dplyr::select(annot.id, symbol)
pleio_tib <- pleio_mixed_list %>%
  purrr::map(.f = function(x) rename(x, gene1_id = local_gene_id, gene2_id = nonlocal_gene_id)) %>%
  c(pleio_local_list, pleio_nonlocal_list) %>%
  bind_rows() %>%
  rename(pleiotropy_lod = lod) %>%
  inner_join(hot_annot, by = c("gene1_id" = "annot.id")) %>%
  rename(gene1_symbol = symbol) %>%
  inner_join(hot_annot, by = c("gene2_id" = "annot.id")) %>%
  rename(gene2_symbol = symbol) %>%
  dplyr::select(gene1_symbol, gene2_symbol, pleiotropy_lod)
```

```{r pm-{{hot_chr}}}
pm <- pleio_tib %>%
  arrange(gene1_symbol, gene2_symbol) %>%
  tibble_to_matrix(symmetric = TRUE)
```

```{r, scatter-{{hot_chr}}, fig.cap="Pleiotropy test statistic vs. mediation LOD difference for Chr {{hot_chr}} hotspot."}
p <- pleio_mixed_list %>%
  bind_rows() %>%
  rename(pleiotropy_lod = lod) %>%
  inner_join(med_tib, by = c("local_gene_id", "nonlocal_gene_id")) %>%
  dplyr::select(local_symbol, nonlocal_symbol, pleiotropy_lod, lod_diff, lod_diff_proportion, lod_without_med, lod_with_med, filename, nonlocal_gene_id, local_gene_id) %>%
  ggplot() + geom_point(mapping = aes(x = pleiotropy_lod, y = lod_diff, colour = local_symbol == "{{keller_mediator}}", nonlocal = nonlocal_symbol, local = local_symbol), size = 0.05) + theme(legend.position = "none") + xlab("Pleiotropy LOD") + ylab("Mediation LOD difference")
plotly::ggplotly(p)
```

```{r, scatter-by-local-{{hot_chr}}, fig.height = ceiling(nlocal / 2), fig.cap="Pleiotropy test statistic vs. mediation LOD difference by local expression trait for Chr {{hot_chr}} hotspot."}
foo <- pleio_mixed_list %>%
  bind_rows() %>%
  rename(pleiotropy_lod = lod) %>%
  inner_join(med_tib, by = c("local_gene_id", "nonlocal_gene_id")) %>%
  dplyr::select(local_symbol, nonlocal_symbol, pleiotropy_lod, lod_diff, lod_diff_proportion, lod_without_med, lod_with_med, filename, nonlocal_gene_id, local_gene_id)
p <- foo %>%
  ggplot() +
  geom_point(mapping = aes(x = pleiotropy_lod, y = lod_diff, colour = nonlocal_symbol), size = 0.1) +
  facet_wrap(. ~ local_symbol, nrow = ceiling(nlocal / 2), ncol = 2) +
  broman::karl_theme(strip.placement = "inside") +
  theme(legend.position = "none") +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank()
) +
  xlab("Pleiotropy LOD") +
  ylab("Mediation LOD difference") +
  geom_text(aes(x = max(foo$pleiotropy_lod), y = 1, label = local_symbol), inherit.aes=FALSE)
plotly::ggplotly(p)
```

```{r, scols-{{hot_chr}}}
scols1 <- tibble(local = 1L + (rownames(pm) %in% local_annot2$symbol), 
                 symbol = rownames(pm), 
                 mediated_by_keller_mediator = rownames(pm) %in% (med_tib %>% filter(local_symbol == "{{keller_mediator}}", lod_diff > 1.5))$nonlocal_symbol
                 )
```

```{r, ordering-{{hot_chr}}}
# order the nonlocal traits first
## indicator for being a nonlocal trait
indic_nonlocal <- rownames(pm) %in% nonlocal_annot$symbol
# define ord_nonlocal
ord_nonlocal <- pm[indic_nonlocal, ] %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  dendextend::seriate_dendrogram(dist(pm[indic_nonlocal, ])) %>%
  order.dendrogram()
rn_nonlocal <- rownames(pm[indic_nonlocal, ][ord_nonlocal, ])
# define ord_local
ord_local<- pm[!indic_nonlocal, ] %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  dendextend::seriate_dendrogram(dist(pm[!indic_nonlocal, ])) %>%
  order.dendrogram()
rn_local <- rownames(pm[!indic_nonlocal, ][ord_local, ])
rn <- c(rn_nonlocal, rn_local) # order for dendrogram
```

@@

```{r allele-effects-{{hot_chr}}}
fn <- here::here("analysis", "data", "derived_data", "Chr{{hot_chr}}_scan1coef_out.rds")
if (file.exists(fn)) {
  s1coef_out <- readRDS(fn)
} else {
s1coef_out <- purrr::map(.x = as_tibble(cbind(hot_local, hot_nonlocal)), 
           .f = function(pheno) {
             ph <- as.matrix(pheno)
             rownames(ph) <- rownames(expr)
             qtl2::scan1coef(genoprobs = list(probs[["{{hot_chr}}"]]), 
                             pheno = ph, 
                             kinship = K[["{{hot_chr}}"]], 
                             addcovar = covar, 
                             reml = TRUE, 
                             cores = 1
                             )
           }
                )
saveRDS(s1coef_out, fn)
}
```

```{r, get-allele-effects-{{hot_chr}}}
# add index values for peak position
local_nonlocal_annot <- dplyr::bind_rows(local_annot2, nonlocal_annot) %>%
  mutate(peak_index = match(table = names(map[["{{hot_chr}}"]]), x = marker.id)) %>%
  mutate(peak_effects = purrr::map2(.f = qtl2pleio::get_effects, .x = peak_index, .y = s1coef_out, map = map[["{{hot_chr}}"]])) %>%
  qtl2hotspots::listcol_to_cols(list_column = quo(peak_effects), another_column = quo(symbol)) %>%
  rename(cA = `1`, cB = `2`, cC = `3`, cD = `4`, cE = `5`, cF = `6`, cG = `7`, cH = `8`
         )
```

```{r, n-med-{{hot_chr}}}
n_med_tr <- 10 # arbitrary threshold for number of mediated traits per local
# only those local traits with n_med_tr or more mediated nonlocal traits 
# appear as annotation columns in the heatmap
n_med_per_local <- med_tib %>%
  group_by(local_symbol) %>%
  filter(lod_diff > 1.5) %>%
  count() %>%
  filter(n >= n_med_tr)
n_med_per_local %>%
  arrange(desc(n)) %>%
  knitr::kable(caption = "Number of mediated traits per local trait for Chr {{hot_chr}} hotspot.")
```

```{r, prepare-annotations-{{hot_chr}}}
# need 5 annotation object: 1. local v nonlocal, 2. mediation LOD difference for all local traits, 3. mediation LOD difference proportion for all local traits, 4. 8 allele effects per trait, 5. peak position per trait, as a one-column heatmap
# 2,3,4 are numeric; 1 & 5 are binary
## local_nonlocal_annot already has the allele effect and local v nonlocal; binning has the peak position. Need to join these two on symbol
### med_tib has lod diff & lod diff proportion for all local - nonlocal pairs
fake_tibble_mediation <- tidyr::crossing(symbol = local_annot2$symbol, 
                                         symbol2 = local_annot2$symbol) %>%
  mutate(dat = NA) %>%
  spread(key = symbol, value = dat) %>%
  rename(symbol = symbol2)
# to add to mediation tibbles

# first med tibble
lod_diffs_tibble <- med_tib %>%
  select( - c(lod_without_med, 
              lod_with_med, 
              lod_diff_proportion, 
              local_gene_id, 
              nonlocal_gene_id)
          ) %>%
  tidyr::spread(key = local_symbol, 
                value = lod_diff) %>% 
  dplyr::rename(symbol = nonlocal_symbol) %>%
  dplyr::bind_rows(fake_tibble_mediation)
# second med tibble
lod_diff_proportions_tibble <- med_tib %>%
  select( - c(lod_without_med, 
              lod_with_med, 
              lod_diff, 
              local_gene_id, 
              nonlocal_gene_id)
          ) %>%
  tidyr::spread(key = local_symbol, 
                value = lod_diff_proportion) %>% # 147 rows by 32 columns
  rename(symbol = nonlocal_symbol) %>%
  bind_rows(fake_tibble_mediation)
# effects tibble
allele_effects_tibble <- local_nonlocal_annot %>%
  dplyr::select(symbol, 18:25)
# peak positions tibble
peak_positions <- local_nonlocal_annot %>%
  select(symbol, pos)
# local indicator
local_tibble <- local_nonlocal_annot %>%
  select(symbol, local_tx) %>% 
  ungroup()
# univariate lods
lod_tibble <- local_nonlocal_annot %>%
  select(symbol, lod) %>%
  ungroup()
```

```{r, order-annotations-{{hot_chr}}}
# all annots 
annots_ord <- list(lod_diffs_tibble, 
                   lod_diff_proportions_tibble, 
                   allele_effects_tibble,
                   peak_positions, 
                   local_tibble,
                   lod_tibble 
                    ) %>% 
  purrr::map(.f = function(df, rn){
    out <- df[match(x = rn, table = df$symbol), ]
    return(out)
  }, 
  rn = rn)
```

```{r, heat-{{hot_chr}}, fig.cap="Heatmap for pleiotropy test statistics for Chr {{hot_chr}}."}
ord <- match(x = rn, table = rownames(pm))

pm[ord, ord] %>% # order both cols and rows by 'ord'
  apply(FUN = trunc2, MARGIN = 2, threshold = 5) %>%
  pheatmap::pheatmap(cluster_rows = FALSE,
                     cluster_cols = FALSE,
                     fontsize = 2
                     )
```

```{r, index-v-position-{{hot_chr}}, fig.cap="QTL peak position vs. dendrogram index for Chr {{hot_chr}}."}
bar <- foo %>%
  filter(local_symbol == "{{keller_mediator}}")

p <- tibble(nonlocal_symbol = rownames(pm[indic_nonlocal, ])[ord_nonlocal], index = 1:length(nonlocal_symbol)) %>%
  inner_join(bar, by = "nonlocal_symbol") %>% 
  left_join(nonlocal_annot, by = c("nonlocal_symbol" = "symbol")) %>%
  ggplot() + geom_point(aes(y = pos, x = index, color = lod_diff > 1.5, nonlocal = nonlocal_symbol, lod_diff = lod_diff))

plotly::ggplotly(p)
```

```{r {{hot_chr}}-ann3}
ann3 <- annots_ord[[3]] %>%
  ungroup() %>%
  select(- symbol) %>%
  as.matrix() %>%
  (function(x){rownames(x) <- annots_ord[[3]]$symbol; return(x)})
colnames(ann3) <- LETTERS[1:8]
```

```{r {{hot_chr}}-ann2}
ann2 <- annots_ord[[2]] %>%
  select(-symbol) %>%
  as.matrix() %>%
  (function(x){rownames(x) <- annots_ord[[2]]$symbol; return(x)})
```

```{r {{hot_chr}}-ann1}
ann1 <- annots_ord[[1]] %>%
  select(-symbol) %>%
  as.matrix() %>%
  (function(x){rownames(x) <- annots_ord[[1]]$symbol; return(x)})
```

```{r {{hot_chr}}-ann5}
ann5 <- annots_ord[[5]] %>%
  select(-symbol) %>%
  as.matrix() %>%
  (function(x){rownames(x) <- annots_ord[[5]]$symbol; return(x)})
```

```{r {{hot_chr}}-ann6}
ann6 <- annots_ord[[6]] %>%
  select(-symbol) %>%
  as.matrix() %>%
  (function(x){rownames(x) <- annots_ord[[6]]$symbol; return(x)})
```

```{r {{hot_chr}}-ann4}
ann4 <- annots_ord[[4]] %>%
  ungroup() %>%
  select(-symbol) %>%
  as.matrix() %>%
  (function(x){rownames(x) <- annots_ord[[4]]$symbol; return(x)})
```

```{r iheatmapr-{{hot_chr}}, out.width="25in", out.height = "10in"}
mymat <- pm[ord, ord] %>% # order both cols and rows by 'ord'
  apply(FUN = trunc2, MARGIN = 2, threshold = 5)
iheatmapr::main_heatmap(mymat, show_colorbar = FALSE, layout = list(font = list(size = 4))) %>%
  iheatmapr::add_col_labels() %>%
  iheatmapr::add_col_annotation(ann5) %>%
  iheatmapr::add_main_heatmap(ann3, name = "signal3", size = 0.2, show_colorbar = FALSE) %>%
    iheatmapr::add_col_labels() %>%

  iheatmapr::add_main_heatmap(ann2, name = "signal2", size = 0.5, show_colorbar = FALSE) %>%
  iheatmapr::add_col_labels() %>%
  iheatmapr::add_main_heatmap(ann1, name = "signal1", size = 0.5, show_colorbar = FALSE) %>%
  iheatmapr::add_row_annotation(ann5, side = "left") %>%
    iheatmapr::add_col_labels() %>%
  iheatmapr::add_row_annotation(ann4, side = "left") %>%
  iheatmapr::add_row_annotation(ann6, side = "left") %>%
  iheatmapr::add_row_labels()
```

>>>>

@@

```{r, hot_annot-{{hot_chr}}}
hot_annot <- local_annot2 %>%
  bind_rows(nonlocal_annot) %>%
  dplyr::select(annot.id, symbol)
```

@@

### TIMBR analyses

```{r, timbr-setup}
tr <- cbind(hot_local, hot_nonlocal)
prior_M1 <- list(model.type = "crp", # crp - Chinese Restaurant Process
  prior.alpha.type = "gamma",
  prior.alpha.shape = 1,
  prior.alpha.rate = 2.333415)
prior_M2 <- list(model.type = "crp",
                prior.alpha.type = "gamma",
                prior.alpha.shape = 2.3009322,
                prior.alpha.rate = 0.7488104
                )
library(TIMBR)
data(mcv.data) # get A matrix
peaks <- qtl2::find_peaks(s1out, map)
```

```{r}
fn <- "attie_DO500_genoprobs_v5_36state_chr{{hot_chr}}.rds"
gp <- readRDS(here::here("analysis", "data", "derived_data", fn))
```

```{r, timbr}
pheno <- cbind(hot_local, hot_nonlocal)
set.seed(3411192) # to ensure getting the same samples with TIMBR
future::plan(cluster)
t_out <- peaks %>%
  dplyr::filter(chr == hot_chr, pos <= hot_upper, pos >= hot_lower) %>%
  dplyr::mutate(marker_index = purrr::map2_int(.x = pos, .y = chr, .f = function(x, y) which(map[[y]] == x))) %>%
  dplyr::mutate(t_out = furrr::future_pmap(.l = list(lodcolumn, chr, marker_index), .f = function(lodcolumn, chr, marker_index){
    phe_pre <- pheno[ , colnames(pheno) == lodcolumn, drop = FALSE]
    phe <- phe_pre[!is.na(phe_pre), , drop = FALSE]
    pr_pre <- gp[[1]][, , marker_index] # use first (and only!) entry in gp
    pr <- pr_pre[rownames(pr_pre) %in% rownames(phe), ]
    cov_pre <- cbind(1, covar)
    cov_pre2 <- cov_pre[rownames(cov_pre) %in% rownames(phe), ]
    cov <- qtl2:::drop_depcols(cov_pre2)
    TIMBR::TIMBR(y = phe, 
                 prior.D = list(P = pr, 
                                A = mcv.data$prior.D$A, # works for autosomes
                                fixed.diplo = FALSE), 
                 prior.M = prior_M1, 
                 samples = 10000, 
                 samples.ml = 10000,
                 Z = cov, 
                 verbose = TRUE
                 )  
}))
saveRDS(t_out, here::here("analysis", "data", "derived_data", "timbr-chr{{hot_chr}}-10ksamples-M1.rds"))
```

### 2020-03-25 analysis/paper/Rmd/hotspots-paper-timbr.Rmd
set up for using child processes with TIMBR analyses

---
title: "Expression QTL hotspot dissection with mediation analysis and pleiotropy testing"
author: "Frederick J. Boehm, Brian S. Yandell, and Karl W. Broman"
date: "August 3, 2019"
bibliography: research.bib
linkReferences: true
link-citations: true 
output: 
  bookdown::html_document2: 
    code_folding: hide
    self_contained: no
    keep_md: true
  bookdown::pdf_document2:
    fig_caption: true
    keep_tex: true
---

```{r child = "abstract.Rmd", eval = TRUE}
```

```{r child="intro.Rmd", eval = TRUE}
```

@@

```{r setup_continued}
hot_chr <- c(2, 5, 7, 11, 13)
hot_mid <- c(165.5, 146, 46, 71, 112.5)
keller_mediator <- c("Hnf4a", "Pdx1", "Fam83e", "Sat2", "Il6st")
inputs <- tibble(hot_chr, hot_mid, keller_mediator)
out <- pmap_dfc(inputs, .f = function(hot_chr, hot_mid, keller_mediator) knitr::knit_expand(text = readLines(here::here("analysis", "paper", "Rmd", "hotspot-one-chromosome.Rmd"))))
# note that ... argument must be a list, ie, it must be a single argument, not multiple args
```

@@

```{r read_results, child = "results.Rmd", eval = TRUE}
```

```{r read_discussion, child = "discussion.Rmd", eval = TRUE}
```

>>>>

---
title: "Expression QTL hotspot dissection with mediation analysis and pleiotropy testing"
author: "Frederick J. Boehm, Brian S. Yandell, and Karl W. Broman"
date: "March 25, 2020"
bibliography: research.bib
linkReferences: true
link-citations: true 
output: 
  bookdown::html_document2: 
    code_folding: hide
    self_contained: no
    keep_md: true
  bookdown::pdf_document2:
    fig_caption: true
    keep_tex: true
---

```{r setupx, include=FALSE, dev="CairoPNG"}
.libPaths("~/work/libs")
knitr::opts_chunk$set(dev="CairoPNG")
```

@@

```{r setup_continued}
hot_chr <- c(2, 5, 7, 11, 13)
hot_mid <- c(165.5, 146, 46, 71, 112.5)
keller_mediator <- c("Hnf4a", "Pdx1", "Fam83e", "Sat2", "Il6st")
inputs <- tibble(hot_chr, hot_mid, keller_mediator)
out <- pmap_dfc(inputs, .f = function(hot_chr, hot_mid, keller_mediator) knitr::knit_expand(text = readLines(here::here("analysis", "paper", "Rmd", "hotspot-one-chromosome-timbr.Rmd"))))
# note that ... argument must be a list, ie, it must be a single argument, not multiple args
```

@@
